---
title: "ActivityAnalysis"
author: "Daniel Kunk"
date: "`r Sys.Date()`"
output: html_document
---
We can first load any global packages

```{r}
library(tidyverse)
```

Below we will load the activity CSV.

Also, we will prep the dataset for use. And also set some output vars.

```{r}
# Set file path
file_path = "../GenerateActivityBins/diet_1min_activity_downsample_rollingMean.csv"

# Load and prep the analysis dataframe in a single pipeline
df = read_csv(file_path, show_col_types = FALSE) %>%
  mutate(
    # Add day/night categorical labels (night is 16ZT to 0ZT, day is 0 to 16)
    time_of_day = if_else((ZT_day %% 24) < 16, "day", "night"),
    is_photophase = (ZT_day %% 24) < 16,
    
    # Add day number
    day_num = floor(ZT_continuous / 24),
    
    # Convert pixels to cm (281 px/cm)
    distance_cm = filtered_distance / 281.0,
    
    # Create a truly unique identifier for each animal
    animal_id = paste(replicate, "- Ind", individual)
  ) %>%
  # Sort the dataframe
  arrange(treatment, replicate, individual, ZT_continuous)


# Set color palettes for LD and DD
palette_dict = c("LD-diet" = "#EE99AA", "DD-diet" = "#994455")

# Set and create output directories
out_dir = "./Plots/"
tables_dir = "./Tables/"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(tables_dir, showWarnings = FALSE, recursive = TRUE)


# Quickly preview the dataframe
head(df)
```

First we will start with our modelling where we average over all 4 days.

Note that we want to normalize per hour since day and night are different length. 

Additionally, day 1 started recording at 4ZT so its technically only 12 hours...

We need to account for that...


```{r}
daily_dist = df %>%
  group_by(treatment, animal_id, day_num, is_photophase) %>%
  summarise(
    distance_cm = sum(distance_cm, na.rm = TRUE),
    observed_minutes = n(), # Counts the actual number of 1-minute rows recorded
    .groups = "drop"
  ) %>%
  mutate(
    # Convert observed minutes to hours
    duration_hours = observed_minutes / 60,
    
    # Normalize to a rate
    dist_per_hour = distance_cm / duration_hours,
    
    # Create clean grouping variables for modeling and plotting
    phase_label = if_else(is_photophase, "Day", "Night"),
    group = factor(paste(treatment, phase_label, sep = "_"))
  ) %>%
  mutate(treatment = factor(treatment, levels = c("LD-diet", "DD-diet")))
```

Now we have things filtered we can go ahead and start modelling.

```{r}
library(glmmTMB)
library(DHARMa)

# Fit a Tweedie mixed model
tweedie_mod = glmmTMB(
  dist_per_hour ~ group + (1 | animal_id), 
  data = daily_dist, 
  family = tweedie(link = "log")
)

summary(tweedie_mod)

# Simulate residuals and plot DHARMa diagnostics (QQ and Res vs Pred plots)
res_dharma = simulateResiduals(tweedie_mod, n = 500)
plot(res_dharma) 

# Identify potential outliers mathematically
testOutliers(res_dharma)



library(emmeans)

# Extract estimated marginal means on the response scale
emm = emmeans(tweedie_mod, ~ group, type = "response")
emm_simple_summary = summary(emm) %>% 
  as.data.frame() %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))


# Calculate pairwise comparisons with Tukey correction
pw_results = pairs(emm, adjust = "tukey")
pw_df = as.data.frame(summary(pw_results))

summary(pw_results)

# Save to disc
write_csv(as.data.frame(summary(pw_results)), file.path(tables_dir, "pairwise_glmmtmb_tukey_averaged.csv"))

# write estimated marginal means to disc
write_csv(emm_simple_summary, file.path(tables_dir, "emm_glmmtmb_averaged.csv"))

```

# start the ggplot
```{r}
library(ggplot2)
library(stringr)
library(ggbeeswarm)

# Calculate scaling heights before building the brackets dataframe
max_val = max(daily_dist$dist_per_hour, na.rm = TRUE)
max_day = max(daily_dist$dist_per_hour[daily_dist$phase_label == "Day"], na.rm = TRUE)
max_night = max(daily_dist$dist_per_hour[daily_dist$phase_label == "Night"], na.rm = TRUE)

y_day   = max_day + (max_val * 0.05)
y_night = max_night + (max_val * 0.05)
highest_drawn_y = max(y_day, y_night)

# Define the helper functions
get_p = function(g1, g2) {
  pattern = paste0("(", g1, ".*", g2, ")|(", g2, ".*", g1, ")")
  match_row = pw_df %>% filter(str_detect(contrast, pattern))
  if (nrow(match_row) > 0) return(match_row$p.value[1]) else return(1.0)
}

p_to_stars = function(p) {
  case_when(
    p < 0.0001 ~ "****",
    p < 0.001  ~ "***",
    p < 0.01   ~ "**",
    p < 0.05   ~ "*",
    TRUE       ~ "ns"
  )
}

# Clean up x-axis labels for the plot
daily_dist = daily_dist %>%
  mutate(
    phase_label_clean = factor(
      phase_label, 
      levels = c("Day", "Night"),
      labels = c("Day /\nSubjective Day", "Night /\nSubjective Night")
    )
  )

# Calculate bracket coordinates
brackets = tibble(
  comp = c("Day_LD_vs_DD", "Night_LD_vs_DD", "LD_Day_vs_Night", "DD_Day_vs_Night", "Cross_LD_Day_vs_DD_Night"),
  # LD sits at -0.1875 (left), DD sits at +0.1875 (right)
  xmin = c(1 - 0.1875, 2 - 0.1875, 1 - 0.1875, 1 + 0.1875, 1 - 0.1875),
  xmax = c(1 + 0.1875, 2 + 0.1875, 2 - 0.1875, 2 + 0.1875, 2 + 0.1875),
  y_position = c(
    y_day, 
    y_night, 
    highest_drawn_y + (max_val * 0.12), 
    highest_drawn_y + (max_val * 0.20),
    highest_drawn_y + (max_val * 0.28)
  ),
  p_val = c(
    get_p("LD-diet_Day", "DD-diet_Day"),
    get_p("LD-diet_Night", "DD-diet_Night"),
    get_p("LD-diet_Day", "LD-diet_Night"),
    get_p("DD-diet_Day", "DD-diet_Night"),
    get_p("LD-diet_Day", "DD-diet_Night") 
  )
) %>%
  mutate(annotation = p_to_stars(p_val)) %>%
  filter(annotation != "ns")

# Start the final plot
activity_plot = ggplot(daily_dist, aes(x = phase_label_clean, y = dist_per_hour, fill = treatment)) +
  geom_boxplot(
    width = 0.5, 
    position = position_dodge(0.75), 
    outlier.shape = NA, 
    alpha = 0.7, 
    color = "black"
  ) +
  geom_point(
    aes(group = treatment), 
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    color = "black", 
    alpha = 0.4, 
    size = 1.5, 
    stroke = 0
  ) +
  # Use labels argument to simplify the legend output without changing your raw data
  scale_fill_manual(
    values = palette_dict,
    labels = c("LD-diet" = "LD", "DD-diet" = "DD")
  ) +
  labs(
    x = NULL, 
    y = "Distance Travelled (cm / hour)", 
    fill = NULL
  ) +
  #scale_y_continuous(expand = expansion(mult = c(0.02, 0.1))) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    axis.text = element_text(color = "black"),
    axis.line = element_line(color = "black")
  )

# Draw brackets
if (nrow(brackets) > 0) {
  tip = max_val * 0.015 
  
  activity_plot = activity_plot + 
    geom_segment(
      data = brackets, 
      aes(x = xmin, xend = xmax, y = y_position, yend = y_position), 
      color = "black", inherit.aes = FALSE
    ) +
    geom_segment(
      data = brackets, 
      aes(x = xmin, xend = xmin, y = y_position, yend = y_position - tip), 
      color = "black", inherit.aes = FALSE
    ) +
    geom_segment(
      data = brackets, 
      aes(x = xmax, xend = xmax, y = y_position, yend = y_position - tip), 
      color = "black", inherit.aes = FALSE
    ) +
    geom_text(
      data = brackets, 
      aes(x = (xmin + xmax) / 2, y = y_position, label = annotation),
      vjust = 0, nudge_y = tip * 0.2, size = 6, color = "black", inherit.aes = FALSE
    )
}

print(activity_plot)
```

## Starting the large model with each day. 

Note we have dropped the last day since it only contains about 4 hours of data.

```{r}
library(glmmTMB)
library(DHARMa)
library(emmeans)
library(dplyr)
library(readr)
library(stringr)

daily_dist_4d = daily_dist %>%
  mutate(day_factor = factor(day_num)) %>%
  filter(day_num %in% 0:3) %>%
  droplevels()   # drops the now-unused day_factor level "4"


# dispersion allowed to vary by phase (fixes the day/night
# residual deviation), fit on day 0-3 data (day 4 dropped)
full_mod3 = glmmTMB(
  dist_per_hour ~ treatment * phase_label * day_factor + (1 | animal_id),
  dispformula = ~ phase_label,
  data = daily_dist_4d,
  family = tweedie(link = "log")
)

summary(full_mod3)

res3 = simulateResiduals(full_mod3, n = 500)
plot(res3)
testOutliers(res3)
testQuantiles(res3)

# Estimated marginal means and pairwise contrasts, computed within each day.
# Tukey adjustment here is applied separately per day_factor level (family size = 4
# groups -> 6 comparisons per day), not across all 4 days.
emm_by_day = emmeans(full_mod3, ~ treatment * phase_label | day_factor, type = "response")
emm_by_day_summary = summary(emm_by_day) %>% 
  as.data.frame() %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))
pw_by_day  = pairs(emm_by_day, adjust = "tukey")
pw_df      = as.data.frame(pw_by_day)

# Classify each of the 6-per-day contrasts into the 4 comparisons we actually
# care about (LD vs DD within a phase; day vs night within a treatment).
# The other 2 per day (the "cross" comparisons, e.g. LD-Day vs DD-Night)
# are kept in the saved table but not plotted.
classify_contrast = function(x) {
  ld_day   = str_detect(x, fixed("LD-diet Day"))
  ld_night = str_detect(x, fixed("LD-diet Night"))
  dd_day   = str_detect(x, fixed("DD-diet Day"))
  dd_night = str_detect(x, fixed("DD-diet Night"))
  case_when(
    ld_day & dd_day     ~ "Day: LD vs DD",
    ld_night & dd_night ~ "Night: LD vs DD",
    ld_day & ld_night    ~ "LD: Day vs Night",
    dd_day & dd_night    ~ "DD: Day vs Night",
    TRUE ~ NA_character_
  )
}

pw_df = pw_df %>% mutate(comparison_type = classify_contrast(contrast))

# save the full per-day table (all 6 comparisons x 4 days) for the record
write_csv(pw_df, file.path(tables_dir, "pairwise_glmmtmb_within_day.csv"))

# write the emmans table as well
write_csv(emm_by_day_summary, file.path(tables_dir, "emm_glmmtmb_within_day.csv"))

```


```{r}
library(ggplot2)
library(purrr)
library(tibble)
library(dplyr)

p_to_stars = function(p) {
  case_when(
    p < 0.0001 ~ "****",
    p < 0.001  ~ "***",
    p < 0.01   ~ "**",
    p < 0.05   ~ "*",
    TRUE       ~ "ns"
  )
}

daily_dist_4d = daily_dist_4d %>%
  mutate(
    treatment = factor(treatment, levels = c("LD-diet", "DD-diet")),
    phase_label_clean = factor(
      phase_label,
      levels = c("Day", "Night"),
      labels = c("Day /\nSubjective Day", "Night /\nSubjective Night")
    )
  )

global_max = max(daily_dist_4d$dist_per_hour, na.rm = TRUE)

# build the 4 annotation brackets for a single day, using that day's own
# box heights but margins scaled off the global max (keeps spacing
# consistent across panels)
build_day_brackets = function(day_lvl) {
  d = daily_dist_4d %>% filter(day_factor == day_lvl)

  get_max = function(trt, phase) {
    vals = d$dist_per_hour[d$treatment == trt & d$phase_label == phase]
    if (length(vals) == 0) return(0)
    max(vals, na.rm = TRUE)
  }
  get_p = function(type) {
    row = pw_df %>% filter(day_factor == day_lvl, comparison_type == type)
    if (nrow(row) == 0) return(1)
    row$p.value[1]
  }

  y_day   = max(get_max("LD-diet", "Day"),   get_max("DD-diet", "Day"))   + global_max * 0.05
  y_night = max(get_max("LD-diet", "Night"), get_max("DD-diet", "Night")) + global_max * 0.05
  base    = max(y_day, y_night)

  tibble(
    day_factor      = day_lvl,
    comparison_type = c("Day: LD vs DD", "Night: LD vs DD", "LD: Day vs Night", "DD: Day vs Night"),
    xmin        = c(1 - 0.1875, 2 - 0.1875, 1 - 0.1875, 1 + 0.1875),
    xmax        = c(1 + 0.1875, 2 + 0.1875, 2 - 0.1875, 2 + 0.1875),
    y_position  = c(y_day, y_night, base + global_max * 0.12, base + global_max * 0.20),
    p_val       = c(get_p("Day: LD vs DD"), get_p("Night: LD vs DD"),
                     get_p("LD: Day vs Night"), get_p("DD: Day vs Night"))
  )
}

brackets_all = map_dfr(levels(daily_dist_4d$day_factor), build_day_brackets) %>%
  mutate(annotation = p_to_stars(p_val)) %>%
  filter(annotation != "ns")

tip = global_max * 0.015

# set brackets
bracket_paths = brackets_all %>%
  mutate(bracket_id = row_number()) %>%
  pmap_dfr(function(xmin, xmax, y_position, day_factor, bracket_id, ...) {
    tibble(
      bracket_id = bracket_id,
      day_factor = day_factor,
      x = c(xmin, xmin, xmax, xmax),
      y = c(y_position - tip, y_position, y_position, y_position - tip)
    )
  })

activity_plot = ggplot(daily_dist_4d, aes(x = phase_label_clean, y = dist_per_hour, fill = treatment)) +
  geom_boxplot(width = 0.5, position = position_dodge(0.75), outlier.shape = NA, alpha = 0.7, color = "black") +
  geom_point(
    aes(group = treatment),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    color = "black", alpha = 0.4, size = 1.2, stroke = 0
  ) +
  scale_fill_manual(values = palette_dict, labels = c("LD-diet" = "LD", "DD-diet" = "DD")) +
  labs(x = NULL, y = "Distance Travelled (cm / hour)", fill = NULL) +
  facet_wrap(~ day_factor, nrow = 2, labeller = as_labeller(function(x) paste0("Day ", as.integer(x) + 1))) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "top", legend.title = element_blank(),
    axis.text = element_text(color = "black"), axis.line = element_line(color = "black")
  )

if (nrow(brackets_all) > 0) {
  activity_plot = activity_plot +
    geom_path(
      data = bracket_paths,
      aes(x = x, y = y, group = bracket_id),
      inherit.aes = FALSE, color = "black",
      lineend = "butt", linejoin = "mitre"
    ) +
    geom_text(
      data = brackets_all,
      aes(x = (xmin + xmax) / 2, y = y_position, label = annotation),
      inherit.aes = FALSE, vjust = 0, nudge_y = tip * 0.2, size = 5, color = "black"
    )
}

print(activity_plot)
ggsave(file.path(out_dir, "distance_rate_glmmTMB_byDay_pubReady.svg"),
       plot = activity_plot, width = 6, height = 6)
```

